---
title: Introduction to Generalized Additive Models
author: "Nick Giangreco"
output: html_document
---

```{r setup,message=F,warning=F}
knitr::opts_chunk$set(echo = TRUE,dpi = 400)
```

# Motorcycle crash data: linear approach


```{r}

mcycle <- MASS::mcycle

# Examine the mcycle data frame
head(mcycle)
plot(mcycle$times,mcycle$accel)

# Fit a linear model
lm_mod <- lm(accel ~ times, data = mcycle)

# Visualize the model
termplot(lm_mod, partial.resid = TRUE, se = TRUE)

```

# Motorcycle crash data: non-linear approach

```{r}

library(mgcv)

# Fit the model
gam_mod <- gam(accel ~ s(times), data = mcycle)

# Plot the results
plot(gam_mod, residuals = TRUE, pch = 1)

```

# Parts of a non-linear function

```{r}

coef(gam_mod)

```

The smooth is made up of 9 basis functions, each with their own coefficient. A GAM with just two variables can have many coefficients, which means they require a bit more data than linear models.

# Setting complexity of the motorcycle model

```{r}

# Fit a GAM with 3 basis functions
gam_mod_k3 <- gam(accel ~ s(times, k = 3), data = mcycle)

# Fit with 20 basis functions
gam_mod_k20 <- gam(accel ~ s(times, k = 20), data = mcycle)

# Visualize the GAMs
par(mfrow = c(1, 2))
plot(gam_mod_k3, residuals = TRUE, pch = 1)
plot(gam_mod_k20, residuals = TRUE, pch = 1)


```

# Using smooth parameters to avoid overffitting

```{r}

gam_mod <- gam(accel ~ s(times), data = mcycle, method = "REML")
gam_mod$sp

# Fix the smoothing parameter at 0.1
gam_mod_s1 <- gam(accel ~ s(times), data = mcycle, sp = 0.1)

# Fix the smoothing parameter at 0.0001
gam_mod_s2 <- gam(accel ~ s(times), data = mcycle, sp =0.0001)

# Plot both models
par(mfrow = c(2, 1))
plot(gam_mod_s1, residuals = TRUE, pch = 1)
plot(gam_mod_s2, residuals = TRUE, pch = 1)

```

# Complexity and smoothing together

```{r}

# Fit the GAM
gam_mod_sk <- gam(accel ~ s(times, k=50), data=mcycle, sp=0.0001)

# Visualize the model
plot(gam_mod_sk, residuals = TRUE, pch = 1)


```

# Multivariate GAMs of auto performance

```{r}

library(gamair)
data("mpg", package="gamair")

# Examine the data
head(mpg)
str(mpg)

# Fit the model
mod_city <- gam(city.mpg ~ s(weight) + s(length) + s(price), 
                data = mpg, method = "REML")

# Plot the model
plot(mod_city, pages = 1)

```

# Adding categorical to the auto performance plot

```{r}

mod_city2 <- gam(city.mpg ~ s(weight) + s(length) + s(price) + fuel + drive + style, 
                 data = mpg, method = "REML")

# Plot the model
plot(mod_city2, all.terms = TRUE, pages = 1)

```

# Category-level smooths for different auto types

```{r}

# Fit the model
mod_city3 <- gam(city.mpg ~ s(weight, by=drive) + s(length, by=drive) + s(price, by=drive) + drive, 
                 data = mpg, method = "REML")

# Plot the model
plot(mod_city3, pages = 1)


```